---
title: "collegehelper Package Demo"
author: "Suvan Giri"
date: "2025-12-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load the package (use ONE of these methods):
# Method 1: If package is installed
# library(collegehelper)

# Method 2: If package isn't installed yet (recommended for now)
devtools::load_all()

# Method 3: Emergency backup - define functions here
# hw_mean <- function(x) mean(x, na.rm = TRUE)
# (add other functions if needed)
```

## Introduction to R Markdown

This document demonstrates the **collegehelper** package that I created for my R Programming final project. R Markdown is a really useful tool because it lets you combine regular text, executable R code, and mathematical notation all in one document. When you click the **Knit** button, R Studio runs all the code chunks and generates a nicely formatted HTML page with the results embedded.

R Markdown is way better than copying and pasting results into Word because everything stays reproducible - if the data changes, you just re-knit and all the numbers and graphs update automatically.

### Why R Markdown Matters

One of the coolest things about R Markdown is that you can include mathematical formulas using LaTeX syntax. For example, here's an inline equation: $y = \alpha + \beta x + \epsilon$. This is the classic linear regression formula.

You can also display equations on their own line like this:

$$
\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i
$$

This is the formula for the arithmetic mean, which is exactly what my `hw_mean()` function calculates!

## About the collegehelper Package

The **collegehelper** package provides simple functions to help college students analyze their homework scores. It includes three main functions:

1. `hw_mean()` - calculates the mean while handling missing values
2. `hw_stats()` - creates an S3 object with comprehensive statistics
3. `hw_hist()` - generates a histogram visualization

Let me show you how each function works with some example data.

## Creating Sample Data

First, let's create a vector of homework scores. I'm including an `NA` value to demonstrate how the functions handle missing data:

```{r create-data}
# Sample homework scores from a hypothetical semester
scores <- c(80, 90, 75, 100, NA, 85, 92, 78, 88, 95)

# Display the scores
print(scores)
```

## Function 1: Basic Mean Calculation

The `hw_mean()` function calculates the average score while automatically removing any missing values. This is super helpful because you don't have to remember to add `na.rm = TRUE` every time:

```{r mean-demo}
# Calculate the mean score
average_score <- hw_mean(scores)
print(paste("Average homework score:", round(average_score, 2)))
```

This function also includes defensive programming - it checks that the input is numeric before trying to calculate the mean. If you pass in text or other non-numeric data, it will give a clear error message instead of crashing.

## Function 2: Comprehensive Statistics (S3 Object)

The `hw_stats()` function is more advanced. It returns an S3 object that contains multiple statistics about your homework scores:

```{r stats-demo}
# Get comprehensive statistics
stats_result <- hw_stats(scores)

# Display the statistics using the custom print method
print(stats_result)
```

This demonstrates the S3 class system - the `hw_stats` object has a custom `print.hw_stats()` method that displays the information in a clean, readable format instead of just showing a raw list.

Let's access some individual statistics:

```{r access-stats}
# Access specific statistics from the S3 object
cat("The standard deviation is:", round(stats_result$sd, 2), "\n")
cat("Score range:", stats_result$min, "to", stats_result$max, "\n")
cat("Total assignments:", stats_result$n, "\n")
cat("Missing assignments:", stats_result$n_missing, "\n")
```

## Function 3: Histogram Visualization

Finally, the `hw_hist()` function creates a histogram to visualize the distribution of scores. This helps you quickly see patterns like whether most scores are clustered together or spread out:

```{r histogram-demo, fig.width=7, fig.height=5}
# Create a histogram with 8 bins
hw_hist(scores, bins = 8)
```

The histogram shows that most of my scores are in the 80-95 range, which is pretty good! You can adjust the number of bins to change how detailed the visualization is:

```{r histogram-detailed, fig.width=7, fig.height=5}
# Create a more detailed histogram with 10 bins
hw_hist(scores, bins = 10)
```

## Defensive Programming Demo

One important aspect of this package is defensive programming. Here's what happens when you try to pass invalid input to the functions:

```{r error-demo, error=TRUE}
# Try to calculate mean of text (this will show an error)
bad_input <- c("A", "B", "C")
hw_mean(bad_input)
```

Instead of crashing with a confusing error, the function gives you a clear message about what went wrong. This is much more helpful for users!

## Additional Examples

Let's test the functions with some different scenarios:

```{r additional-examples}
# Example 1: Perfect scores (no missing data)
perfect_scores <- c(95, 98, 100, 97, 99, 100)
cat("Perfect semester average:", hw_mean(perfect_scores), "\n\n")

# Example 2: More missing data
incomplete_scores <- c(80, NA, 75, NA, 85, 90, NA)
incomplete_stats <- hw_stats(incomplete_scores)
print(incomplete_stats)
```

```{r viz-comparison, fig.width=7, fig.height=5}
# Visualize the incomplete scores
hw_hist(incomplete_scores, bins = 6)
```

## Comparing My Package to Base R

Here's why this package is useful compared to writing everything manually:

```{r comparison}
# Without collegehelper (the old, tedious way):
raw_scores <- c(80, 90, 75, 100, NA, 85)
manual_mean <- mean(raw_scores, na.rm = TRUE)
manual_sd <- sd(raw_scores, na.rm = TRUE)
manual_n <- sum(!is.na(raw_scores))
cat("Manual calculation:\n")
cat("Mean:", manual_mean, "\n")
cat("SD:", manual_sd, "\n")
cat("N:", manual_n, "\n\n")

# With collegehelper (much simpler!):
cat("Using collegehelper:\n")
hw_stats(raw_scores)
```

The package saves time and reduces the chance of making mistakes!

## Final Thoughts

This package demonstrates several key concepts from the R Programming course:

### Module 10: DESCRIPTION File
- Complete DESCRIPTION file with all required metadata
- Proper version numbering (0.0.0.9000 for development)
- Author information and package description
- Dependency specification (only base R needed)
- CC0 license for open educational use

### Module 11: Debugging & Defensive Programming
- Input validation with clear error messages
- Handling of missing values (NA)
- Type checking to prevent crashes
- Fixed the Tukey outlier bug (using `&` instead of `&&` for element-wise operations)

### Module 12: R Markdown Documentation
- Mix of text, code, and mathematical notation
- LaTeX equations: $\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i$
- Multiple code chunks with real outputs
- Data visualizations embedded in the document
- Reproducible analysis workflow

### Object-Oriented Programming
- S3 class system (`hw_stats` class)
- Custom print method (`print.hw_stats()`)
- Clean, organized output structure

### Package Features Summary
- **3 unique functions**: `hw_mean()`, `hw_stats()`, `hw_hist()`
- **0 external dependencies**: Uses only base R
- **S3 object system**: Custom class with print method
- **Defensive programming**: Input validation throughout
- **Complete documentation**: Roxygen2 help files for all functions

This makes collegehelper a clean, production-ready student package that actually solves a real problem - tracking homework performance throughout the semester!

## Source Code

All source code is available on GitHub at: https://github.com/SuviG36/collegehelper.git

---

*Document generated on `r Sys.Date()` using R Markdown. The complete package includes this vignette, function documentation, and all source code.*
